# Testing and Estimation - Large vs. Small Samples

This lecture note explores the concepts of testing and estimation in the context of large and small samples.


## Central Limit Theorem (CLT)

The Central Limit Theorem (CLT) is a fundamental concept in statistics that describes the behavior of sample means when drawing repeated samples from a population. It states that, regardless of the shape of the population distribution, the distribution of sample means will tend to be approximately normal as the sample size increases.

## Conceptual Explanation

The CLT is based on three key principles:

1. **Sampling Distribution**: When we draw multiple samples from a population and calculate the mean of each sample, we create a sampling distribution of the sample means.

2. **Approximate Normality**: The distribution of sample means will be approximately normal if the sample size is sufficiently large, regardless of the shape of the population distribution. This is particularly true for sample sizes greater than 30.

3. **Mean and Standard Deviation**: The mean of the sample means will be approximately equal to the population mean, and the standard deviation of the sample means (standard error) will decrease as the sample size increases.

**Example 1**
Suppose we want to understand the distribution of the mean height of students in the campus of Gadau. We'll collect the heights of 10000 students as our population data.

```{r gather_data, echo=TRUE}
# Simulate student heights in Gadau campus
set.seed(123)
student_heights <- rnorm(1000, mean = 170, sd = 10)

# Display summary statistics of student heights
summary(student_heights)
# Number of samples
num_samples <- 1000
# Sample size
sample_size <- 30

# Initialize a vector to store sample means
sample_means <- numeric(num_samples)

# Take random samples and calculate means
for (i in 1:num_samples) {
  sample <- sample(student_heights, sample_size)
  sample_means[i] <- mean(sample)
}

# Plot the distribution of sample means
hist(sample_means, breaks = 30, main = "Distribution of Sample Means", xlab = "Sample Mean Height (cm)", col = "lightblue", border = "white")

```
In the histogram above, we can observe that as the number of samples increases, the distribution of sample means becomes more symmetric and bell-shaped, resembling a normal distribution. This is consistent with the Central Limit Theorem.

**Example 2: Rolling a Fair Die**

Suppose we want to understand the distribution of the mean of rolling a fair six-sided die. We will simulate rolling the die 1000 times and calculate the mean for each set of rolls.

```{r example_code, echo=TRUE}
# Set seed for reproducibility
set.seed(123)

# Simulate rolling a fair six-sided die 1000 times
num_rolls <- 1000
die_rolls <- sample(1:6, num_rolls, replace = TRUE)

# Calculate the mean of each set of rolls
sample_means <- numeric(num_rolls)
for (i in 1:num_rolls) {
  sample_means[i] <- mean(sample(die_rolls, i))
}

# Plot the distribution of sample means
hist(sample_means, breaks = 30, main = "Distribution of Sample Means", xlab = "Sample Mean", col = "lightblue", border = "white")
```


